---
title: "Bubble Sheet Marker"
subtitle: "BEES2041 Part 1 MCQ grading via Claude vision API"
format: html
---


Amazing

- bubble locations still not quite right
  - scanned pdfs have been printed and image size shrubk slightly
- empty bubbles getting marked as filled
- function to update marks and reprint bubble layout preview
- seperate code into functions


Grades scanned BEES2041 Part 1 MCQ bubble sheets in four steps:

1. **Calibrate** — extract precise bubble positions from the blank PDF (once, before first use)
2. **Preprocess** — convert scanned PDF pages to JPEG images and initialise a progress CSV
3. **Mark** — call Claude vision API for each page, updating the CSV as results arrive
4. **Score** — compare detected answers against the answer key and produce `results.csv`

## Prerequisites

Install R packages:

```{r, eval=FALSE}
install.packages(c("httr2", "pdftools", "jsonlite", "readr", "dplyr",
                   "tibble", "base64enc", "optparse", "magick"))
```

`pdftools` requires the system library **poppler**:

| OS | Command |
|----|---------|
| macOS | `brew install poppler` |
| Ubuntu/Debian | `sudo apt install libpoppler-cpp-dev` |

Set your Anthropic API key (add to `~/.zshrc` to make permanent):

```{bash}
export ANTHROPIC_API_KEY="sk-ant-..."
```

## Workflow

**0.** Set working directory:

```{bash}
cd "week 12/2026-BEES2041-Exam"
```

**1.** Calibrate bubble positions from the blank bubble sheet PDF (run once; overwrites `support/bubble_layout.R` and writes `support/bubble_layout_preview.jpeg`):

```{bash}
Rscript support/bubble_calibrate_coords.R
```

Open `support/bubble_layout_preview.jpeg` to verify that the red letters land inside the correct bubbles on the blank form. Re-run after any layout change.

**2.** Convert the scanned PDF to per-page images and initialise `progress.csv`:

```{bash}
Rscript support/bubble_preprocess_pdf.R responses/scans.pdf
```

This creates `responses/` containing `page_NNNN.jpeg` files and `progress.csv`.
It also creates `responses/overrides.csv` for manual scoring overrides.

**3.** Test on the first 3 pages before committing to a full run:

```{bash}
Rscript support/bubble_marker.R --dir responses/ --dry-run
```

Check `responses/marked/` — each page image should show detected answer letters overlaid in red next to the filled bubbles.

**4.** Run the full marking pass (resumable — pages already marked are skipped):

```{bash}
Rscript support/bubble_marker.R --dir responses/
```

During marking, a pixel-based blank-row safeguard is applied after model parsing:

- Rows with very low bubble-center signal are forced to blank.
- Ambiguous weak-dominance rows can also be forced blank when the model letter disagrees with the pixel winner.
- The safeguard is intentionally conservative to avoid blanking genuinely filled rows.
- Any forced blanks are recorded in `notes` as `Pixel blank-check forced blank: ...`.

**5.** Review any rows where `needs_review == TRUE` in `responses/progress.csv` against the physical sheets and correct the relevant `q*` columns manually.

Also review rows where `notes` includes `Pixel blank-check forced blank:` and confirm those question rows are genuinely blank on the paper.

For an audit trail, record manual answer corrections in `responses/overrides.csv` with columns:

- `file` (e.g. `pages/page_0001.jpeg`)
- `zid` (optional if `file` is provided)
- `question` (numeric, e.g. `14`)
- `response` (`A`–`E`, or blank to clear an answer)

Example override row:

```{text}
pages/page_0001.jpeg,z5108583,14,
```

This example clears Q14 for that sheet (forces it to blank for scoring).

**6.** Score against the answer key and produce `results.csv`:

```{bash}
Rscript support/bubble_results.R --dir responses/
```

**7.** Import `responses/results.csv` into the grade spreadsheet. The first four rows are the answer keys (one per version); student rows follow sorted by version then zID.

```{r}
library(tidyverse)


# Merge with master
master <- readxl::read_xlsx("/Users/dfalster/Library/CloudStorage/OneDrive-UNSW/teaching/2026 BEES2041/assessments/4-Exam/marking.xlsx", sheet = 1)

results <- list.files("responses", pattern = "results.csv", full.names = TRUE, recursive = TRUE) |>
  lapply(readr::read_csv) |>
  dplyr::bind_rows()

corrections <- read_csv("responses/zid_corrections.csv")
corrections |> filter(!corrected %in% master$Username)
corrections |> filter(duplicated(corrected))
corrections |> filter(duplicated(zid))

# Apply zid corrections
unmatched_corrections <- corrections |> filter(!zid %in% results$zid)
if (nrow(unmatched_corrections) > 0) {
  message("WARNING: ", nrow(unmatched_corrections), " correction(s) had no matching zid in results:")
  print(unmatched_corrections)
} else {
  message("All corrections matched.")
}

results <- results |>
  left_join(corrections |> select(zid, corrected), by = "zid") |>
  mutate(zid = if_else(!is.na(corrected), corrected, zid)) |>
  select(-corrected)

results |>
  filter(!(zid %in% master$Username)) |>
  write_csv("responses/no_matches.csv")

master |> select(Username) |> left_join(results, by = c("Username" = "zid")) |>
  select(Username, exam_version, page, score, pct) |>
  write_csv("responses/merged_results.csv")

```

## Script options

### bubble_calibrate_coords.R

No options. Reads `output/bubblesheet_v1.pdf` and writes:

- `support/bubble_layout.R` — precise normalised bubble coordinates
- `support/bubble_layout_preview.jpeg` — annotated blank form for visual verification

### bubble_preprocess_pdf.R

| Option | Default | Description |
|--------|---------|-------------|
| `scans.pdf` | *(required)* | Path to scanned bubble sheet PDF (positional) |
| `--dpi` | `150` | JPEG resolution in DPI |
| `--force` | off | Re-extract even if output folder exists |

### bubble_marker.R

| Option | Default | Description |
|--------|---------|-------------|
| `--dir` | *(required)* | Folder created by bubble_preprocess_pdf.R |
| `--model` | `claude-sonnet-4-6` | Anthropic model ID |
| `--dry-run` | off | Process only first 3 pending pages |

### bubble_results.R

| Option | Default | Description |
|--------|---------|-------------|
| `--dir` | *(required)* | Folder containing `progress.csv` |
| `--key` | `output/answer_key.csv` | Answer key CSV |
| `--output` | `<dir>/results.csv` | Output CSV path |

## Output files

**`<dir>/marked/page_NNNN-marked.jpeg`** — annotated copy of every successfully graded page. The detected answer letter for each question is printed in red next to the corresponding bubble. Open this folder in an image viewer for rapid visual QC.

**`<dir>/progress.csv`** — one row per page, updated live during marking:

| Column | Description |
|--------|-------------|
| `file` | JPEG filename |
| `page` | Page number in the original PDF |
| `status` | `"pending"` / `"done"` / `"error"` |
| `api_call_ok` | `TRUE` if the API call succeeded |
| `needs_review` | `TRUE` if uncertain (`*`) answers are present or an error occurred |
| `zid` | Student zID (e.g. `z1234567`) |
| `exam_version` | Exam version read from sheet (1–4) |
| `confidence` | `"high"` or `"low"` as reported by Claude |
| `notes` | Claude notes plus any pixel blank-check actions |
| `error` | Error message if `status == "error"` |
| `q1` … `q26` | Detected letter (A–E) per question (25 columns; Q16 and Q27 absent) |

**`<dir>/overrides.csv`** — optional manual corrections used by `bubble_results.R` before scoring:

| Column | Description |
|--------|-------------|
| `file` | Page file path (e.g. `pages/page_0001.jpeg`) |
| `zid` | Student zID (optional if `file` is provided) |
| `question` | Question number (1–15, 17–26) |
| `response` | Override answer (`A`–`E`) or blank to clear |

**`<dir>/results.csv`** — scored output produced by `bubble_results.R`:

| Column | Description |
|--------|-------------|
| `zid` | Student zID, or `KEY_v1`–`KEY_v4` for answer key rows |
| `exam_version` | Exam version (1–4) |
| `score` | Total correct out of 25 (NA for key rows) |
| `pct` | Percentage score (NA for key rows) |
| `q1` … `q26` | `1` = correct, `0` = incorrect (25 columns) |

## Scan quality notes

- **Resolution:** 150 DPI minimum; 200 DPI safer for faintly-filled sheets.
- **Orientation:** Scan portrait with auto-rotate off — the script reads "Version X" from the top-right corner.
- **Skew:** Keep < 2°; heavy skew causes misreads.
- **Contrast:** Normal or slightly high — faint bubbles are the main source of low-confidence flags.
- **Double-filled bubbles:** Claude flags these as low confidence; spot-check the physical sheet.
- **Single PDF:** Merge all pages before running (`pdfunite page*.pdf scans.pdf` on Linux; macOS Preview → Print → Save as PDF).

## Cost estimate

Using `claude-sonnet-4-6` at 150 DPI (~400 KB/page): ~170 students costs **$0.50–$1.50** per run.
Always test with `--dry-run` first.
